TAPIR: Threaded Application Programming on Intermittent Resources

### 2017-05-11 paper.tex
Added section for interrupt implementation

\section{System Implementation } % TODO change title
Mutexes - Emily\\

TAPIR has a round robin scheduling algorithm. At the end of a task, the compiler
inserts a jump back to the scheduler task, where the next task in the list is
the one to be executed. An alternate scheduling model that was considered was
a context switch on reboot instead of at the end of each task. The main drawback
of using a reboot-based scheduler is that, because power failure is the common
case, there is a high chance of reducing or halting forward progress altogether
because one thread holds a lock that all of the other threads need but cannot
complete. For that reason the design specifies end of task, so each task has
an opportunity to use a shared resource and release any locks before a power-
intensive context switch occurs. Additionally, a round robin scheduler was
chosen for this design because it is simple to implement and uses only one index counter,
very little power or memory overhead as compared to other scheduling algorithms. It is
critical for multithreading in Chain to not add significant power and memory
overhead in scheduling over conventional Chain so that the workload can
make as much forward progress as possible on one charge cycle.

@@

    // Index of current running thread in threads[]
    // Stored in self channel
    unsigned current;
\end{lstlisting}
\caption{Scheduler Data Structures}\label{schDataStruct}
\end{minipage}\hfill
\end{figure*}

@@

Mutexes (Simple, meant for pseudo channels) - Emily\\
Interrupts (mucked with IE flags) - Neil\\
% Interrupt prologue
% Interrupt setup
%\section{Implementation}
%Scheduler Data structures and idempotence (what we broke and why) -NEIL\\
%Mutexes (Simple, meant for pseudo channels) - Emily\\
%Interrupts (mucked with IE flags) - Neil\\
%% Interrupt prologue
%% Interrupt setup

>>>>

\section{System Implementation } % TODO change title
TAPIR has a round robin scheduling algorithm. At the end of a task, the compiler
inserts a jump back to the scheduler task, where the next task in the list is
the one to be executed. An alternate scheduling model that was considered was
a context switch on reboot instead of at the end of each task. The main drawback
of using a reboot-based scheduler is that, because power failure is the common
case, there is a high chance of reducing or halting forward progress altogether
because one thread holds a lock that all of the other threads need but cannot
complete. For that reason the design specifies end of task, so each task has
an opportunity to use a shared resource and release any locks before a power-
intensive context switch occurs. Additionally, a round robin scheduler was
chosen for this design because it is simple to implement and uses only one index counter,
very little power or memory overhead as compared to other scheduling algorithms. It is
critical for multithreading in Chain to not add significant power and memory
overhead in scheduling over conventional Chain so that the workload can
make as much forward progress as possible on one charge cycle.

@@

    // Index of current running thread
    // Stored in self channel
    unsigned current;
\end{lstlisting}
\caption{Scheduler Data Structures}\label{schDataStruct}
\end{minipage}\hfill
\end{figure*}

@@

\subsection{Interrupts}
The primary issue with interrupts is the interrupt enable flags. The MSP430,
like many other embedded arhitetures has a master interrupt enable,
as well as specific, device-level interrupt enables. Both are the user's
responsibility, and need to be specified in the user-defined
\texttt{INIT\_FUNC}, as both will be cleared on reset. When an interrupt fires,
however, interrupts will be disabled by the CPU (again, on most architectures)
and if the device loses power immediately after an interrupt occurs, there will
be no record that the interrupt occured. This case is unaviodable on
intermittent power without explicit support from the power system - we reason
that if the user cared about every interrupt that fired and if missing one
interrupt would be catestrophic to the device's operation, then, arguably, that
device should not be intermittently powered.

Beyond this, there remains the issue of how to return to the interrupt task
with the interrupt enable set to the correct value if a power failure occurs.
Since interrupts are disabled on boot (on the MSP430, for instance), the user
needs to add code to their \texttt{INIT\_FUNC} to enable interrupts. However,
since interrupt handlers are assumed to be non-reentrant on our system, if
we are inside an interrupt handler, we need to keep interrupts disabled. We
also need to write the address of the task that we should return to upon boot -
the address of the \texttt{INTERRUPT\_TASK} on our system. If these are two
separate variables, there is a correctness issue with respect to power
failures. If we write the location of the \texttt{INTERRUPT\_TASK}, but die
before we can write the fact that we are in an interrupt handler, we will
return to te interrupt handler with interrupts enabled (since they will be
enabled by the \texttt{INIT\_TASK}. If we reverse the order and write the fact
that we are in an interrupt handler, then die before we can write the address
of the \texttt{INTERRUPT\_TASK}, we will resume execution at a non-interrupt
task with interrupts disabled, since the \texttt{INIT\_TASK} will not re-enable
them. Both of these cases could cause execution to fail; the former fails if
the interrupt handler is not written to be reentrant, the latter fails since
interrupts will never be processed. Thus, these writes need to be atomic.

We could depend on the programmer to designate all tasks that would be run as
part of the interrupt handler, then use whether the current task address was
equivalent to one of those as the flag that designates that we are in an
interrupt handler, but this forces tasks to either be used as a part of the
interrupt handler or not, which is bad for code reuse. This also prevents the
user from being able to spawn off non-interrupt tasks as part of an interrupt
handler.

Instead, we exploit the fact that, generally, architectures are byte-aligned
with a fixed instruction size of 16 bits or more - this means, at the very
least, the lowest bit of any code address is zero. As this is true of ARM,
RISC-V, MSP430, and most RISC architecutres, we feel that this is a reasonable
assumption. We use this fact to store the flag that is set if execution is
inside an interrupt handler in the lowest bit of the address of the task that
we are currently executing (the task that execution will return to upon
reboot). Thus, we can use a single write to encode all the needed information,
giving atomicity. For simplicity for the programmer, upon an interrupt firing,
an \texttt{interrupt\_prologue()} is run that does this write. The address stored
in the current context is still the address of the user-defined
\texttt{INTERRUPT\_TASK}. In our model, this write is the guarentuee that the
system has recieved the interrupt - if power fails between when the interrupt
fires and this write occurs, the interrupt will be ignored. This could be fixed
by having a special capacitor in the power system to accomidate the write, but
assumptions about the power system, arguably, should not be a part of \chain.

To avoid the pathological stack corruption case described in the motivation for
interrupts, we keep a flag in non-volatile memory that is set if a reboot has
occured while in an interrupt handler and is unset otherwise. This flag is
unset in the afformentioned \texttt{interrupt\_prologue()} and set in
\texttt{main()}. If and only if the flag is unset when
\texttt{return\_from\_interrupt()} is called, the two values pushed on the
stack from an interrupt firing are popped off. Otherwise the stack remains
unchanged, since the values will have been effectively cleared by a power
failure.

As of this writing, interrupts are still being debugged, but the above
reasoning seems, in theory, correct.
% TODO Mutexes (Simple, meant for pseudo channels) - Emily\\
